Formalization, Mechanization and Automation of \\ G\"{o}del's Proof of God's Existence\thanks{This work has been supported by the German Research Foundation under grant BE2501/9-1.}

### 2013-08-20 Abstract/GodProof-Abstract.tex

\newcommand{\imp}{\rightarrow}
\newcommand{\biimp}{\leftrightarrow}
\newcommand{\all}{\forall}
\newcommand{\ex}{\exists}
\newcommand{\seq}{\vdash}
\newcommand{\nec}{\Box} % necessarily
\newcommand{\pos}{\Diamond} % possibly
\newcommand{\ess}[2]{#1 \ \mathit{Ess.} \ #2}
\newcommand{\NE}{\mathit{NE}}

\title{ Formalization, Mechanization and Automation % of Variants
  of G\"{o}del's Proof of God's Existence\thanks{This work has been
    supported by the German Research Foundation under grant
    BE2501/9-1.}  }

@@

Dana Scott's version of G\"odel's proof \cite{ScottNotes} employs the
following axioms (\textbf{A}), definitions (\textbf{D}), corollaries
(\textbf{C}) and theorems (\textbf{T}), and it proceeds in the
following order:\footnote{ (A1), (A2), (A5), (D1), (D3) are logically
  equivalent to, respectively, axioms 5, 2 and 4 and definitions 1 and
  3 in G\"odel's notes \cite{Goedel1970,GoedellNotes}. (A3) was introduced by
  Scott \cite{ScottNotes} and could be derived from G\"odel's axiom 1 and
  (D1) in a logic with infinitary conjunction. (A4) is a weaker form
  of G\"odel's axiom 3. (D2) has an extra conjunct lacking in
  G\"odel's definition 2; this lack is believed to have been an
  oversight by G\"odel \cite{Hazen}. }
% \marginpar{\scriptsize Chris: The reference to Scott's version is kind of confusing, we need to verify this source.}
% \marginpar{\scriptsize Bruno: I think the numbers (A1, A2,..,
% D1,..D3, L1,..,L3,T) should be in the left with a description
% environment, as they were before. I think it looked nicer
% before. Can I change it back?}

\allowdisplaybreaks[1] 
\begin{description}
\item[A1] Either the property or its negation is positive, but not
  both:  \hfill 
  $\all \phi. [P(\neg \phi) \biimp \neg P(\phi)]$
\item[A2] A property is positive if it is necessarily implied by a
  positive property:  \\ \phantom{b} \hfill 
  $\all \phi. \all \psi.[(P(\phi) \wedge \nec \all x.[\phi(x)
  \imp \psi(x)]) \imp P(\psi)]$ 
\item[T1] Positive properties are possibly exemplified. \hfill $\all \varphi. [P(\varphi) \imp \pos \ex x.\varphi(x)]$
\item[D1] A God-like being possesses all positive properties: \hfill
  $G(x) \biimp \forall \phi. [P(\phi) \to \phi(x)]$
\item[A3]  The property of being God-like is positive: \hfill   $P(G)$
\item[C\phantom{1}] Possibly, God exists: \hfill $\pos \ex x. G(x)$
\item[A4]  Positive properties are necessarily positive: \hfill 
  $\all \phi.[P(\phi) \to \Box \; P(\phi)]$
\item[D2] An essence of an individual is a property possessed by it and necessarily implying any of its properties:\\
  \phantom{b} \hfill $\ess{\phi}{x} \biimp \phi(x) \wedge \all
  \psi. (\psi(x) \imp \nec \all y. (\phi(y) \imp \psi(y)))$
\item[T2]  Being God-like is an essence of any
  God-like being: \hfill $\all y.[G(y) \imp \ess{G}{y}]$ 
\item[D3] Necessary existance of an individual is the necessary exemplification of all its essences: \\
  \phantom{b} \hfill $\NE(x) \biimp \all \phi.[\ess{\phi}{x} \imp \nec \ex x.\phi(x)]$
\item[A5] Necessary existence is a positive property: \hfill $P(\NE)$
\item[T3] Necessarily, God exists: \hfill $\nec \ex x. G(x)$ 
\end{description}

% \begin{align}
% & \text{Any property necessarily implied by a positive property is positive.} \notag \\
% & \quad 
% \all \varphi.\all \psi.[(P(\varphi) \wedge \nec \all x.[\varphi(x) \imp \psi(x)]) \imp P(\psi)] \tag{A1} \\
% & \text{A property is positive if and only if its negation is not positive.} \notag \\
% & \quad 
% \all \varphi. [P(\neg \varphi) \biimp \neg P(\varphi)]
% \tag{A2} \\
% & \text{The property of being God-like is positive.} \notag \\
% & \quad 
% P(G) 
% \tag{A3} \\
% & \text{Positive properties are necessarily positive.} \notag \\
% & \quad 
% \all \varphi.[P(\varphi) \to \Box \; P(\varphi)]
% \tag{A4} \\
% & \text{Necessary existence is a positive property.} \notag \\
% & \quad 
% P(E)
% \tag{A5} \\
% & \text{Being God-like is having every positive property.} \notag \\
% & \quad 
% G(x) \biimp \forall \varphi. [P(\varphi) \to \varphi(x)]
% \tag{D1} \\ 
% & \parbox[c][2em][c]{0.8\textwidth}{A property is an essence of an individual if and only if the individual has it and every property  that the individual has is necessarily implied by it.}\notag \\
% & \quad 
% \ess{\varphi}{x} \biimp \varphi(x) \wedge \all \psi. (\psi(x) \imp \nec \all y. (\varphi(y) \imp \psi(y)))
% \tag{D2} \\
% & \parbox[c][2em][c]{0.8\textwidth}{An individual necessarily exists if and only if all its essences are necessarily exemplified.} \notag \\
% & \quad 
% E(x) \biimp \all \varphi.[\ess{\varphi}{x} \imp \nec \ex x.\varphi(x)] 
% \tag{D3} 
% \end{align}

% \noindent
% From these axioms and definitions we then infer:
% \begin{align}
% & \text{Positive properties are possibly exemplified.} \notag \\
% & \quad 
% \all \varphi. [P(\varphi) \imp \pos \ex x.\varphi(x)]
% \tag{L1} \\
% & \text{Possibly, God exists.} \notag \\
% & \quad 
% \pos \ex x. G(x)
% \tag{L2} \\
% & \parbox[c][2em][c]{0.8\textwidth}{The property of being God-like is an essence of any God-like being.} \notag \\
% & \quad 
% \all y.[G(y) \imp \ess{G}{y}]
% \tag{L3} \\
% & \text{Necessarily, God exists.} \notag \\
% & \quad 
% \nec \ex x. G(x)
% \tag{T}
% \end{align}
% \marginpar{\scriptsize Bruno: I think it might be possible to improve the natural language statements further. After our improvements converge here, we should update the statements in all formalizations (THF, Coq, Isabelle, ND).}

The above version of G\"{o}del's proof has now been 
analysed for the first-time
with an unprecedent degree of detail and formality; cf.~\cite{FormalTheologyRepository}. The following has been done (and in this order):
\begin{itemize}
\item A detailed natural deduction proof.
%
\item A formalization of the axioms, definitions and theorems in the TPTP THF syntax \cite{J22}.
%
\item Automatic verification of the consistency of the axioms and 
definitions with Nitpick \cite{Nitpick}.
%
\item Automatic demonstration of the theorems with the provers LEO-II \cite{LEO-II} and Satallax \cite{Satallax}.

@@

In our ongoing computer-assisted study of G\"odel's proof we have
obtained the following results:
\begin{itemize}
\item Logic S5 is not needed for obtaining the results, logic KB is
  sufficient.
\item For proving theorem T1, only the left to right direction of
  axiom A1 is needed. However, the backward direction of A1 is
  required for showing T2.
\item The first conjunct $\phi(x)$ in D2, introduced by Scott, is important; without it the set of axioms and definitions becomes inconsistent.
\end{itemize}

This work attests the maturity of contemporary interactive and
automated deduction tools for classical higher-order logic and it
demonstrates the elegance and practical relevance of the embeddings
based approach.  Most importantly, our work opens new perspectives for
a computer-assisted theoretical philosophy.  The critical discussion
of the underlying concepts, definitions and axioms remains a human
responsibility, but the computer can assist in building and checking
rigorously correct logical arguments. In case of logico-philosophical
disputes, the computer can check the disputing arguments and partially
fulfill Leibniz' dictum: Calculemus --- Let us calculate!

Future work includes a substantial widening of our computer-assisted
study of ontological arguments. Due to the flexibility of the
embeddings based approach, variations of the distinctive features of
the base logic (e.g. non-rigid symbols, varying domains, etc.)  will
be enabled in these studies

>>>>

\newcommand{\imp}{\rightarrow}
\newcommand{\biimp}{\leftrightarrow}
\newcommand{\all}{\forall}
\newcommand{\ex}{\exists}
\newcommand{\seq}{\vdash}
\newcommand{\nec}{\Box} % necessarily
\newcommand{\pos}{\Diamond} % possibly
\newcommand{\ess}[2]{#1 \ \mathit{ess.} \ #2}
\newcommand{\NE}{\mathit{NE}}

\title{ Formalization, Mechanization and Automation % of Variants
  of \\ G\"{o}del's Proof of God's Existence\thanks{This work has been
    supported by the German Research Foundation under grant
    BE2501/9-1.}  }

@@

Dana Scott's version of G\"odel's proof \cite{ScottNotes} employs the
following axioms (\textbf{A}), definitions (\textbf{D}), corollaries
(\textbf{C}) and theorems (\textbf{T}), and it proceeds in the
following order:\footnote{ A1, A2, A5, D1, D3 are logically
  equivalent to, respectively, axioms 5, 2 and 4 and definitions 1 and
  3 in G\"odel's notes \cite{Goedel1970,GoedelNotes}. 
  A3 was introduced by Scott \cite{ScottNotes} and 
  could be derived from G\"odel's axiom 1 and
  D1 in a logic with infinitary conjunction. 
  A4 is a weaker formof G\"odel's axiom 3. 
  D2 has an extra conjunct $\phi(x)$ lacking in G\"odel's definition 2; 
  this is believed to have been 
  an oversight by G\"odel \cite{Hazen}.}

\allowdisplaybreaks[1] 
\begin{description}
\item[A1] Either a property or its negation is positive, but not
  both:  \hfill 
  $\all \phi. [P(\neg \phi) \biimp \neg P(\phi)]$
\item[A2] A property necessarily implied \\ by a
  positive property is positive: \phantom{b} \hfill 
  $\all \phi. \all \psi.[(P(\phi) \wedge \nec \all x.[\phi(x)
  \imp \psi(x)]) \imp P(\psi)]$ 
\item[T1] Positive properties are possibly exemplified: \hfill $\all \varphi. [P(\varphi) \imp \pos \ex x.\varphi(x)]$
\item[D1] A \emph{God-like} being possesses all positive properties: \hfill
  $G(x) \biimp \forall \phi. [P(\phi) \to \phi(x)]$
\item[A3]  The property of being God-like is positive: \hfill   $P(G)$
\item[C\phantom{1}] Possibly, God exists: \hfill $\pos \ex x. G(x)$
\item[A4]  Positive properties are necessarily positive: \hfill 
  $\all \phi.[P(\phi) \to \Box \; P(\phi)]$
\item[D2] An \emph{essence} of an individual is  \\ a property possessed by it and \\ necessarily implying any of its properties:
  \phantom{b} \hfill $\ess{\phi}{x} \biimp \phi(x) \wedge \all
  \psi. (\psi(x) \imp \nec \all y. (\phi(y) \imp \psi(y)))$
\item[T2]  Being God-like is an essence of any
  God-like being: \hfill $\all x.[G(x) \imp \ess{G}{x}]$ 
\item[D3] \emph{Necessary existence} of an individual is \\ the necessary exemplification of all its essences: 
  \phantom{b} \hfill $\NE(x) \biimp \all \phi.[\ess{\phi}{x} \imp \nec \ex y.\phi(y)]$
\item[A5] Necessary existence is a positive property: \hfill $P(\NE)$
\item[T3] Necessarily, God exists: \hfill $\nec \ex x. G(x)$ 
\end{description}

\noindent
The above version of G\"{o}del's proof has now been 
analysed for the first-time
with an unprecedent degree of detail 
and formality; cf.~\cite{FormalTheologyRepository}. 
The following has been done (and in this order):
\begin{itemize}
\item A detailed natural deduction proof.
%
\item A formalization of the axioms, definitions and theorems in the TPTP THF syntax \cite{J22}.
%
\item Automatic verification of the consistency of the axioms and 
definitions with Nitpick \cite{Nitpick}.
%
\item Automatic demonstration of the theorems with the provers LEO-II \cite{LEO-II} and Satallax \cite{Satallax}.

@@

In our ongoing computer-assisted study of G\"odel's proof we have
obtained the following results:
\begin{itemize}
\item The basic modal logic K is sufficient for proving T1, C and T2. \marginpar{\scriptsize{Bruno: check this.}}
\item Modal logic S5 is not needed for proving T3; the logic KB is
  sufficient. \marginpar{\scriptsize{Bruno: Maybe cite a reference for S5 and KB?}}
\item Without the first conjunct $\phi(x)$ in D2 the set of axioms 
  and definitions would be inconsistent.
\item For proving theorem T1, only the left to right direction of
  axiom A1 is needed. However, the backward direction of A1 is
  required for proving T2.

\end{itemize}

This work attests the maturity of contemporary interactive and
automated deduction tools for classical higher-order logic and
demonstrates the elegance and practical relevance of the embeddings-based approach.  Most importantly, our work opens new perspectives for
a computer-assisted theoretical philosophy.  The critical discussion
of the underlying concepts, definitions and axioms remains a human
responsibility, but the computer can assist in building and checking
rigorously correct logical arguments. In case of logico-philosophical
disputes, the computer can check the disputing arguments and partially
fulfill Leibniz' dictum: Calculemus --- Let us calculate!

\marginpar{\scriptsize{Bruno: I find this last paragraph a bit vague and not as self-contained as the rest of the abstract. We could try to improve it, but I think simply removing it would also be satisfactory. The sentence about Leibniz' dictum already provides a very nice ``climax'' for the end of the paper. Anything that we say after it feels like an ``anticlimax''...}}
Future work shall widen our computer-assisted approach to the
study of other ontological arguments. Variations of the distinctive features of
the base logic (e.g. non-rigid symbols, varying domains, etc.), made easier by the flexibility of the
embeddings-based approach, will
be considered.
